Syntax and Semantics of Cedille

### 2020-01-28 docs/semantics/arxiv-version.tex
Update docs to reference Agda version 2.5.4 (ticket #25)
docs: semantics: remove some differences between ArXiV and default version

    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t \simeq t \}}{\Gamma\vdash \textit{FV}(t)\subseteq \textit{dom}(\Gamma)}  \\ \\    
    \infer{\Gamma\vdash \delta\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
    \infer{\Gamma\vdash \rho\ t'\ \mhyph\ t \Leftrightarrow [t_2/x]T}
          {\Gamma\vdash t' \tpsynth t_1 \simeq t_2 & \Gamma \vdash t \Leftrightarrow [t_1/x]T} \\ \\
    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpcheck T'}
          {\Gamma\vdash T\tpcheck \star & \Gamma\vdash t \tpcheck T & T \cong T'} &
    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpsynth T}
          {\Gamma\vdash T\tpcheck \star & \Gamma\vdash t \tpsynth T' & T \cong T'} \\ \\
    \infer{\Gamma\vdash \phi\ t\ \mhyph\ t'\{t''\} \Leftrightarrow T}
          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}  & \
  \end{array}
  \]
\caption{Rules for checking a term against a well-kinded type ($\Gamma \vdash t \tpcheck T$)
           and synthesizing a type for a term ($\Gamma \vdash t \tpsynth T$)}
\label{fig:tp}
\end{figure}

\begin{figure}
  \[
  \begin{array}{lllllll}
    |x| & = & x &\ &
    |\abs{\lambda}{x}{t'}{t}| & = & \absu{\lambda}{x}{|t|} \\
    |t\ t'| & = & |t|\ |t'| &\ &
    |t\cdot T| & = & |t| \\
    |\absu{\Lambda}{x}{t'}{t}| & = & |t| &\ &
    |t\ \mhyph t'| & = & |t| \\
    |[t , t']| & = & |t| &\ &
    |t.1| & = & |t| \\
    |t.2| & = & |t| &\ &
    |\beta\{t\}| & = & |t|\\
    |\delta\ t| & = & |t|&\ &
    |\rho\ t\ \mhyph\ t'| & = & |t'| \\
    |\phi\ t\ \mhyph\ t'\ \{t''\}| & = & |t''| &\ &
    |\chi\ T\ \mhyph\ t'| & = & |t'|
  \end{array}
  \]
  \caption{Erasure for annotated terms}
  \label{fig:eraser}
\end{figure}

@@

Defining \verb|Top| to be $\{\absu{\lambda}{x}{x} \simeq
\absu{\lambda}{x}{x}\}$, we may assign \verb|Top| to any closed term \verb|t|,
including non-normalizing ones.  In our annotated syntax, we write \verb|β{t}|.
Even without this, the presence of
$\delta$ in combination with $\phi$ allows us to type non-normalizing
terms assuming an erased argument $x$ of type $\{ \textit{tt} \simeq
\textit{ff} \}$ for Church-encoded booleans \textit{tt} and
\textit{ff}.  For example, $\delta\ x$ has type
\verb|{λ x . x ≃ λ x . x x}|, and with $\phi$ we can use this to type
$\Omega$ by changing the typed term \verb|id · True id|, where
\verb|True| is \verb|∀ X : ★ . X ➔ X|. But failure of normalization
does not impinge on Theorem~\ref{thm:consis}.   Extensional Martin-L\"of type theory (MLTT) is
also non-normalizing, for a very similar reason, but fact
does not contradict its logical soundness~\cite{dybjer16}.  In CDLE, the guarantees one gets about the
behavior of terms are expressed almost entirely in their types.  If
the types are weak, then not much is guaranteed; but stronger types
can guarantee properties like normalization.

Given the lack of normalization, several checks in the typing rules --
for things like $t =_{\beta\eta} t'$ -- are formally undecidable.  In
practice, we simply impose a bound on the number of steps of reduction,
and thus restore formal decidability (we are checking ``typable within
a given budget'').  In practice, the same is done for Coq and Agda,
where type checking is decidable but, in general, infeasible (since one
may write astronomically slow terminating functions).

@@

\bibitem{agda}
\mbox{The Agda development team}.
\newblock {\em Agda}, 2018.
\newblock Version 2.5.3.

>>>>

    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t \simeq t \}}{\Gamma\vdash \textit{FV}(t)\subseteq \textit{dom}(\Gamma)}  \\ \\    
    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
    \infer{\Gamma\vdash \rho\ t'\ \mhyph\ t \Leftrightarrow [t_2/x]T}
          {\Gamma\vdash t' \tpsynth \{ t_1 \simeq t_2 \} & \Gamma \vdash t \Leftrightarrow [t_1/x]T} \\ \\
    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpcheck T'}
          {\Gamma\vdash T\tpcheck \star & \Gamma\vdash t \tpcheck T & T \cong T'} &
    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpsynth T}
          {\Gamma\vdash T\tpcheck \star & \Gamma\vdash t \tpsynth T' & T \cong T'} \\ \\
    \infer{\Gamma\vdash \phi\ t\ \mhyph\ t'\{t''\} \Leftrightarrow T}
          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}  & \
  \end{array}
  \]
\caption{Rules for checking a term against a well-kinded type ($\Gamma \vdash t \tpcheck T$)
           and synthesizing a type for a term ($\Gamma \vdash t \tpsynth T$)}
\label{fig:tp}
\end{figure}

\begin{figure}
  \[
  \begin{array}{lllllll}
    |x| & = & x &\ &
    |\abs{\lambda}{x}{t'}{t}| & = & \absu{\lambda}{x}{|t|} \\
    |t\ t'| & = & |t|\ |t'| &\ &
    |t\cdot T| & = & |t| \\
    |\abs{\Lambda}{x}{t'}{t}| & = & |t| &\ &
    |t\ \mhyph t'| & = & |t| \\
    |[t , t']| & = & |t| &\ &
    |t.1| & = & |t| \\
    |t.2| & = & |t| &\ &
    |\beta\{t\}| & = & |t|\\
    |\delta\ \mhyph\ t| & = & |t|&\ &
    |\rho\ t\ \mhyph\ t'| & = & |t'| \\
    |\phi\ t\ \mhyph\ t'\ \{t''\}| & = & |t''| &\ &
    |\chi\ T\ \mhyph\ t'| & = & |t'|
  \end{array}
  \]
  \caption{Erasure for annotated terms}
  \label{fig:eraser}
\end{figure}

@@

Defining \verb|Top| to be $\{\absu{\lambda}{x}{x} \simeq
\absu{\lambda}{x}{x}\}$, we may assign \verb|Top| to any closed term \verb|t|,
including non-normalizing ones.  In our annotated syntax, we write \verb|β{t}|.
Even without this, the presence of
$\delta$ in combination with $\phi$ allows us to type non-normalizing
terms assuming an erased argument $x$ of type $\{ \textit{tt} \simeq
\textit{ff} \}$ for Church-encoded booleans \textit{tt} and
\textit{ff}.  For example, $\delta\ \mhyph x$ has type
\verb|{λ x . x ≃ λ x . x x}|, and with $\phi$ we can use this to type
$\Omega$ by changing the typed term \verb|id · True id|, where
\verb|True| is \verb|∀ X : ★ . X ➔ X|. But failure of normalization
does not impinge on Theorem~\ref{thm:consis}.   Extensional Martin-L\"of type theory (MLTT) is
also non-normalizing, for a very similar reason, but this fact
does not contradict its logical soundness~\cite{dybjer16}.  In CDLE, the guarantees one gets about the
behavior of terms are expressed almost entirely in their types.  If
the types are weak, then not much is guaranteed; but stronger types
can guarantee properties like normalization, as demonstrated by the following
theorem:

\begin{theorem}[Call-by-name normalization of functions]
  \label{thm:cedille-termination}
  Suppose \(\Gamma\vdash t : T\), $t$ is closed, and there exists a closed term
  $t'$ which erases to \(\lambda x.\ x\) and whose type is $T \to \Pi x : T_1.\
  T_2$ for some $T_1, T_2$. Then $|t|$ is call-by-name normalizing.
  \end{theorem}

Given the lack of normalization in general, several checks in the typing rules --
for things like $t =_{\beta\eta} t'$ -- are formally undecidable.  In
practice, we simply impose a bound on the number of steps of reduction,
and thus restore formal decidability (we are checking ``typable within
a given budget'').  In practice, the same is done for Coq and Agda,
where type checking is decidable but, in general, infeasible (since one
may write astronomically slow terminating functions).

@@

\bibitem{agda}
\mbox{The Agda development team}.
\newblock {\em Agda}, 2018.
\newblock Version 2.5.4.
